Follow-up Samples

### 2017-01-06 submission/manuscript_outline_20161024.Rmd
fix typo

Colorectal cancer (CRC) continues to be a leading cause of cancer related deaths and is the second most common cancer death among men aged 40-79 years of age [@jemal_cancer_2010; @haggar_colorectal_2009].  Over the last few years death due to the disease has seen a significant decrease thanks mainly to improvements in screening [@jemal_cancer_2010].  However, despite this giant improvement there are still approximately 50,000 deaths from the disease a year [@haggar_colorectal_2009].  It is estimated that around 5-10% of all CRCs can be explained by autosomal dominant inheritance [@green_very_2007]. The vast majority of CRCs are not inherited and the exact etiology to disease has not been well worked out [@haggar_colorectal_2009].  Although many risk factors have been identified [@hagger_colorectal_2009] and non-invasive screening techniques have started to be put into consistent use [@liao_application_2013; @johnson_multi-target_2016] there has been a consistent increase in the incidence of CRC in the younger population.

@@

***Most Important Variables to the Model***
Overall, there were a total of `r format(length(rownames(imp_model_vars)))` variables identified as being present in more than 50% of the training models [Table S2]. The top 5 most important bacterial OTUs were `r paste(rf_lowest_ID[1], " (", names(rf_lowest_ID)[1], ")", sep = "")`,  `r paste(rf_lowest_ID[2], " (", names(rf_lowest_ID)[2], ")", sep = "")`, `r paste(rf_lowest_ID[3], " (", names(rf_lowest_ID)[3], ")", sep = "")`, `r paste(rf_lowest_ID[4], " (", names(rf_lowest_ID)[4], ")", sep = "")`, and `r paste(rf_lowest_ID[5], " (", names(rf_lowest_ID)[5], ")", sep = "")`.  These 5 OTUs were present in at least `r min(filter(imp_model_vars, Variable %in% names(rf_lowest_ID)[1:5])[, "total_appearance"])` out of the total `r length(rownames(model_train_summary))` different 80/20 runs.

@@

There was `r length(rownames(filter(good_metaf, Disease_Free == "n")))` individual who still clearly had CRC on follow up as well as `r length(rownames(filter(good_metaf, Disease_Free == "unknown")))` individuals whose status on follow up was unknown.  Although the `r length(rownames(filter(good_metaf, Disease_Free == "n")))` individual had a decrease in positive probability their follow up sample was still higher than the cutoff threshold of 0.5 (positive probability = `r format(filter(follow_up_prob_summary, sampleType == "followup" & disease_free == "n")[, "Yes"], digits = 3)`).  Interestingly, `r length(rownames(filter(follow_up_prob_summary, sampleType == "followup" & disease_free == "unknown" & Yes > 0.5)))` individuals who were unknown on follow up still were over the threshold cutoff of 0.5 even though, like the `r length(rownames(filter(good_metaf, Disease_Free == "n")))` individual with clear CRC on follow up, the probability of an adenoma or carcinoma decreased [Table S3].

@@

***Specific OTUs in the Lesion Model are not Detected in Follow Up Versus Initial Samples*** 
Overall, there were a total of `r length(rownames(pvalue_common_imp_vars))` OTUs that were common between the main lesion model and the model for classifying initial and follow up samples specifically [Table S4].  A total of `r length(rownames(filter(pvalue_common_imp_vars, BH_corr < 0.05)))` OTU was still significant after multiple comparison correction and it's lowest taxonomic identification was to `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]`.  In general, `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "otu"]` (`r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]`) was decreased from initial to follow up [Figure 5].  The relative abundance was not drastically different then the mean of the values observed in the control training set [Figure 5].

***Differences in Adenoma and Carcinoma in Previously Associated Cancer Bacteria***
First, there was a clear magnitude difference in these specific OTUs based on whether they were from adenoma or carcinoma individuals [Figure 6].  The carcinoma samples showed a significant difference between initial and follow up samples for Peptostreptococcus stomatis (P-value = `r format(pvalue_adn_crc_summary["pept", "crc_BH"], digits = 3)`) and Porphyromonas asaccharolytica (P-value = `r format(pvalue_adn_crc_summary["porp", "crc_BH"], digits = 3)`) whereas there were no signficant differences in any of these OTUs in the adenoma samples [Table S5].

@@

We then investigated which OTUs could potentially be more important in our model [Figure 5 & Table S4].  Many of the OTUs identified classified to normal flora bacterium [Table S4].  Only a single OTU though was significant after multiple comparison correction and the lowest taxonomic identification of `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "otu"]` was to `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]`. Although there was a difference in the relative abundance at initial and follow up these values were not drastically different from the relative abundance values observed in the control individuals of the training set [Figure 5].  Although we were interested in what we could use to classify those with either adenoma or carcinoma versus normal.  We found that the traditional bacteria associated with CRC were higher in magnitude in the carcinoma group and there were significant differences in some of these OTUs between the initial and follow up samples [Figure 5 and Tabl S5].  This research provides evidence that it is possible to use bacterial microbiome data to create a highily sensitive model, that is reactive to therapy, for detection of adenoma or carcinoma.  It accomplishes this by using a unique sample set in which before and after surgery stool samples are available for assessment.  By using these types of samples we are not only able to show sensitivity of lesion prediction but also able to show that this model is reactive.  That is to say that after surgery for removal of the adenoma or carcinoma it decreases the positive probability to reflect a lower likelihood of the individual having an adenoma or carcinoma.

@@

***Study Design and Patient Sampling***
The sampling and design of the study was similar to that reported in Baxter, et al [@baxter_microbiota-based_2016].  In brief, study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline stool sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or Familial adenomatous polyposis.  Samples used to build the model used for prediction were collected either prior to a colonoscopy or between 1 - 2 weeks after.  The bacterial microbiome has been shown to nomralize within this time period (**insert citation**).  Kept apart from this training set were a total of `r length(rownames(good_metaf))` individuals that not only had a sample as described previoulsy but also a follow up sample between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after surgery and treatment had been completed. This study was approved by the University of Michigan Institutional Review Board.  All study participants provided informed consent and the study itself conformed to the guidelines set out by the Helsinki Declaration.

***Fecal Immunochemical Test and 16S rRNA Gene Sequencing***
FIT was analyzed as previously published using both OC FIT-CHEK and OC-Auto Micro 80 automated system (Polymedco Inc.) (**insert citation**). 16S rRNA gene sequencing was completed as previously described by Kozich, et al. [@kozich_development_2013].  In brief, DNA extraction used the 96 well Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075 automated pipetting system (Eppendorf).  The V4 variable region was amplified and the resulting product was split between three sequencing runs with control, adenoma, and carcinoma evenly represented on each run.  Each group was randomly assigned to avoid biases based on sample collection location.

***Sequence Processing***
The mothur software package (v1.37.5) was used to process the 16S rRNA gene sequences.  This process has been previously described (**insert citations**).  The general processing workflow using mothur is as follows: Paired-end reads were first merged into contigs, quality filtered, aligned to the SILVA database, screening for chimeras, classified with a naive Bayesian classifier using the Ribosomal Database Project (RDP), and clustered into Operational Taxonomic Units (OTUs) using a 97% similarity cutoff with an average neighbor clustering algorithm.  The numer of sequences for each sample was rarified to `r ifelse(sd(total_sub_Seqs) == 0, paste(mean(total_sub_Seqs)), paste("Error Check Subsampling"))` in an attempt to minimize uneven sampling.

@@

***Statistical Analysis***
The R software package (v`r paste(version$major, ".", version$minor, sep = "")`) was used for all statisitical analysis.  Comparisons between bacterial community structure utilized PERMANOVA (**insert citation**) in the vegan package (v`r paste(packageVersion("vegan"))`) while comparisons between ROC curves utilized the method by DeLong et al. (**insert citation**) executed by the pROC (v`r paste(packageVersion("pROC"))`) package.  Comparisons between probabilities as well as overall amount of OTU between initial and follow up samples utilized a paired wilcoxson ranked sum test.  Where multiple comparison testing was needed a Benjamini-Hochberg (BH) correction was applied (**insert citation**) and a corrected P-value of less than 0.05 was considered significant.  Unless otherwise stated the P-values reported are those of the BH corrected ones.

***Analysis Overview***
Differences in FIT between initial and follow ups for either adenoma or carcinoma were investigate.  Next, initial and follow up samples were analyzed for differences in alpha and beta diversity.  All OTUs used in the lesion model were also analyzed using a paired wilcoxson test.  The lesion model was then tested for accuracy in prediction and whether it reduced the positive probability of lesion after surgery.  The most important OTUs for this were used to build an updated model and this reduced feature model was assessed for it's similarity to the original model.  We then used the initial follow up model to assess whether this model could classify samples better then the lesion model.  The most important OTUs were then identified from this model and used to create a reduced feature initial follow up model.  This reduced feature model, as was done with the lesion model, was compared to the full model for loss of accuracy.  Finally, in order to investigate the relative abundance of specific bacteria, that have been previously associated with CRC, we selected OTUs that taxonomically classified to Fusobacterium Nucleatum, Parvimonas Micra, Peptostreptococcus Assacharolytica, and Porphyromonas Stomatis.  Specifically, we wanted to test if there were any differences based on whether the individual had an adenoma or carcinoma.

@@

**Figure S1: Thetayc Graphed Against Time of Follow up Sample from Initial**

>>>>

Colorectal cancer (CRC) continues to be a leading cause of cancer related deaths and is the second most common cancer death among men aged 40-79 years of age [@jemal_cancer_2010; @haggar_colorectal_2009].  Over the last few years death due to the disease has seen a significant decrease thanks mainly to improvements in screening [@jemal_cancer_2010].  However, despite this giant improvement there are still approximately 50,000 deaths from the disease a year [@haggar_colorectal_2009].  It is estimated that around 5-10% of all CRCs can be explained by autosomal dominant inheritance [@green_very_2007]. The vast majority of CRCs are not inherited and the exact etiology to disease has not been well worked out [@haggar_colorectal_2009].  Although many risk factors have been identified [@haggar_colorectal_2009] and non-invasive screening techniques have started to be put into consistent use [@liao_application_2013; @johnson_multi-target_2016] there has been a consistent increase in the incidence of CRC in the younger population.

@@

***Most Important Variables to the Model***
Overall, there were a total of `r format(length(rownames(imp_model_vars)))` variables identified as being present in more than 50% of the training models [Figure S2]. The top 5 most important bacterial OTUs were `r paste(rf_lowest_ID[1], " (", names(rf_lowest_ID)[1], ")", sep = "")`,  `r paste(rf_lowest_ID[2], " (", names(rf_lowest_ID)[2], ")", sep = "")`, `r paste(rf_lowest_ID[3], " (", names(rf_lowest_ID)[3], ")", sep = "")`, `r paste(rf_lowest_ID[4], " (", names(rf_lowest_ID)[4], ")", sep = "")`, and `r paste(rf_lowest_ID[5], " (", names(rf_lowest_ID)[5], ")", sep = "")`.  These 5 OTUs were present in at least `r min(filter(imp_model_vars, Variable %in% names(rf_lowest_ID)[1:5])[, "total_appearance"])` out of the total `r length(rownames(model_train_summary))` different 80/20 runs.

@@

There was `r length(rownames(filter(good_metaf, Disease_Free == "n")))` individual who still clearly had CRC on follow up as well as `r length(rownames(filter(good_metaf, Disease_Free == "unknown")))` individuals whose status on follow up was unknown.  Although the `r length(rownames(filter(good_metaf, Disease_Free == "n")))` individual had a decrease in positive probability their follow up sample was still higher than the cutoff threshold of 0.5 (positive probability = `r format(filter(follow_up_prob_summary, sampleType == "followup" & disease_free == "n")[, "Yes"], digits = 3)`).  Interestingly, `r length(rownames(filter(follow_up_prob_summary, sampleType == "followup" & disease_free == "unknown" & Yes > 0.5)))` individuals who were unknown on follow up still were over the threshold cutoff of 0.5 even though, like the `r length(rownames(filter(good_metaf, Disease_Free == "n")))` individual with clear CRC on follow up, the probability of an adenoma or carcinoma decreased [Table S2].

@@

***Specific OTUs in the Lesion Model are not Detected in Follow Up Versus Initial Samples*** 
Overall, there were a total of `r length(rownames(pvalue_common_imp_vars))` OTUs that were common between the main lesion model and the model for classifying initial and follow up samples specifically [Table S3].  A total of `r length(rownames(filter(pvalue_common_imp_vars, BH_corr < 0.05)))` OTU was still significant after multiple comparison correction and it's lowest taxonomic identification was to `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]`.  In general, `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "otu"]` (`r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]`) was decreased from initial to follow up [Figure 5].  The relative abundance was not drastically different then the mean of the values observed in the control training set [Figure 5].

***Differences in Adenoma and Carcinoma in Previously Associated Cancer Bacteria***
First, there was a clear magnitude difference in these specific OTUs based on whether they were from adenoma or carcinoma individuals [Figure 6].  The carcinoma samples showed a significant difference between initial and follow up samples for Peptostreptococcus stomatis (P-value = `r format(pvalue_adn_crc_summary["pept", "crc_BH"], digits = 3)`) and Porphyromonas asaccharolytica (P-value = `r format(pvalue_adn_crc_summary["porp", "crc_BH"], digits = 3)`) whereas there were no signficant differences in any of these OTUs in the adenoma samples [Table S4].

@@

We then investigated which OTUs could potentially be more important in our model [Figure 5 & Table S3].  Many of the OTUs identified classified to normal flora bacterium [Table S3].  Only a single OTU though was significant after multiple comparison correction and the lowest taxonomic identification of `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "otu"]` was to `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]`. Although there was a difference in the relative abundance at initial and follow up these values were not drastically different from the relative abundance values observed in the control individuals of the training set [Figure 5].  Although we were interested in what we could use to classify those with either adenoma or carcinoma versus normal.  We found that the traditional bacteria associated with CRC were higher in magnitude in the carcinoma group and there were significant differences in some of these OTUs between the initial and follow up samples [Figure 5 and Tabl S5].  This research provides evidence that it is possible to use bacterial microbiome data to create a highily sensitive model, that is reactive to therapy, for detection of adenoma or carcinoma.  It accomplishes this by using a unique sample set in which before and after surgery stool samples are available for assessment.  By using these types of samples we are not only able to show sensitivity of lesion prediction but also able to show that this model is reactive.  That is to say that after surgery for removal of the adenoma or carcinoma it decreases the positive probability to reflect a lower likelihood of the individual having an adenoma or carcinoma.

@@

***Study Design and Patient Sampling***
The sampling and design of the study was similar to that reported in Baxter, et al [@baxter_microbiota-based_2016].  In brief, study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline stool sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or Familial adenomatous polyposis.  Samples used to build the model used for prediction were collected either prior to a colonoscopy or between 1 - 2 weeks after.  The bacterial microbiome has been shown to nomralize within this time period [@obrien_impact_2013].  Kept apart from this training set were a total of `r length(rownames(good_metaf))` individuals that not only had a sample as described previoulsy but also a follow up sample between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after surgery and treatment had been completed. This study was approved by the University of Michigan Institutional Review Board.  All study participants provided informed consent and the study itself conformed to the guidelines set out by the Helsinki Declaration.

***Fecal Immunochemical Test and 16S rRNA Gene Sequencing***
FIT was analyzed as previously published using both OC FIT-CHEK and OC-Auto Micro 80 automated system (Polymedco Inc.) [@daly_evaluation_2013]. 16S rRNA gene sequencing was completed as previously described by Kozich, et al. [@kozich_development_2013].  In brief, DNA extraction used the 96 well Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075 automated pipetting system (Eppendorf).  The V4 variable region was amplified and the resulting product was split between three sequencing runs with control, adenoma, and carcinoma evenly represented on each run.  Each group was randomly assigned to avoid biases based on sample collection location.

***Sequence Processing***
The mothur software package (v1.37.5) was used to process the 16S rRNA gene sequences.  This process has been previously described [@kozich_development_2013].  The general processing workflow using mothur is as follows: Paired-end reads were first merged into contigs, quality filtered, aligned to the SILVA database, screening for chimeras, classified with a naive Bayesian classifier using the Ribosomal Database Project (RDP), and clustered into Operational Taxonomic Units (OTUs) using a 97% similarity cutoff with an average neighbor clustering algorithm.  The numer of sequences for each sample was rarified to `r ifelse(sd(total_sub_Seqs) == 0, paste(mean(total_sub_Seqs)), paste("Error Check Subsampling"))` in an attempt to minimize uneven sampling.

@@

***Statistical Analysis***
The R software package (v`r paste(version$major, ".", version$minor, sep = "")`) was used for all statisitical analysis.  Comparisons between bacterial community structure utilized PERMANOVA [@anderson_permanova_2013] in the vegan package (v`r paste(packageVersion("vegan"))`) while comparisons between ROC curves utilized the method by DeLong et al. [@delong_comparing_1988] executed by the pROC (v`r paste(packageVersion("pROC"))`) package.  Comparisons between probabilities as well as overall amount of OTU between initial and follow up samples utilized a paired wilcoxson ranked sum test.  Where multiple comparison testing was needed a Benjamini-Hochberg (BH) correction was applied [@benjamini_controlling_1995] and a corrected P-value of less than 0.05 was considered significant.  Unless otherwise stated the P-values reported are those of the BH corrected ones.

***Analysis Overview***
Differences in FIT between initial and follow ups for either adenoma or carcinoma were investigated.  Next, initial and follow up samples were analyzed for differences in alpha and beta diversity.  All OTUs used in the lesion model were also analyzed using a paired wilcoxson test.  The lesion model was then tested for accuracy in prediction and whether it reduced the positive probability of lesion after surgery.  The most important OTUs for this were used to build an updated model and this reduced feature model was assessed for it's similarity to the original model.  We then used the initial follow up model to assess whether this model could classify samples better then the lesion model.  The most important OTUs were then identified from this model and used to create a reduced feature initial follow up model.  This reduced feature model, as was done with the lesion model, was compared to the full model for loss of accuracy.  Finally, in order to investigate the relative abundance of specific bacteria, that have been previously associated with CRC, we selected OTUs that taxonomically classified to Fusobacterium Nucleatum, Parvimonas Micra, Peptostreptococcus Assacharolytica, and Porphyromonas Stomatis.  Specifically, we wanted to test if there were any differences based on whether the individual had an adenoma or carcinoma.

@@

**Figure S1: Distribution of P-values from Paired Wilcoxson Analysis of OTUs in Initial versus Follow Up**

**Figure S2: Summary of Important Variables in both Lesion and Initial/Follow Up models** A) MDA of the most important variables in the lesion model.  The black point represents the median and the different colors are the different runs up to 100.  B) The total number of appearances of each variable in the 100 different lesion models.  The cutoff of 50% was used to assess importance.

**Figure S3: Thetayc Graphed Against Time of Follow up Sample from Initial**

### 2017-01-06 submission/supplemental_outline_20161024.Rmd

**Table S2: Summary of Most Important Variables For the Model**

```{r supplemental_table2, echo=FALSE, message=FALSE, warning=FALSE}

imp_model_vars <- read.csv('../results/tables/rf_wCV_imp_vars_summary.csv', stringsAsFactors = F, header = T)
rf_otu_tax <- read.csv('../results/tables/rf_otu_tax.csv', stringsAsFactors = F, header = T, row.names = 1)
rf_otu_tax <- cbind(otu = rownames(rf_otu_tax), rf_otu_tax)

all_data <- full_join(imp_model_vars, rf_otu_tax, by = c("Variable" = "otu"))

imp_model_vars <- cbind(Variable = imp_model_vars$Variable, Tax = createTaxaLabeller(all_data), 
                        total_appearance = imp_model_vars$total_appearance)

colnames(imp_model_vars) <- c("Variable", "Lowest Taxonomic ID", "Total Model Appearances")

kable(imp_model_vars, format = "markdown", align = 'c', padding = 2)

**Table S3: Summary of Probability Assignment of Follow Up Individuals**

@@

**Table S4: Summary of OTUs in both Lesion and Initial/Follow Up Models**

@@

**Table S5: Summary of P-values for Specific CRC OTUs between Initial and Follow Ups**

>>>>

**Table S2: Summary of Probability Assignment of Follow Up Individuals**

@@

**Table S3: Summary of OTUs in both Lesion and Initial/Follow Up Models**

@@

**Table S4: Summary of P-values for Specific CRC OTUs between Initial and Follow Ups**
